Lifestyle Choice Matters: Exploring Population Health Disparities and the Effect of Tobacco and Alcohol on Life Expectancy

### 2023-11-03 outputs/paper/paper.qmd
Correct errors in grammar and references

---
title: "Lifestyle Choice Matters: Exploring Population Health Disparities and the Effect of Tobacco and Alcohol on Life Expectancy"
author: "Yufei Liu"
thanks: "Code and data are available at: https://github.com/Florence-Liu/life_expectancy"
date: today
date-format: long
abstract: "Life expectancy is an important measurement of population health and there are many significant influencing factors such as genetic and lifestyle. This paper uses a linear regression model to analyze life expectancy, prevalence of tobacco products, and alcohol consumption data from Global Health Observatory data repository of WHO. We find as prevalence of tobacco increases, the life expectancy at age 60 decreases while as alcohol consumption increases, the life expectancy increases. We also find regional, temporal, and sexual differences in life expectancy at age 60. It implies health disparies in several aspects and encourages the importance of health lifestyle including take less tobacco use and moderate drinking that contribute to increased life expectancy. "
format: pdf
number-sections: true
toc: false
bibliography: references.bib
---

@@

In this paper, we investigate how lifestyle factors influence life expectancy at age 60, and focus on two specific aspects, prevalence of tobacco products and alcohol consumption. We use `R` [@citeR] to analyze life expectancy, prevalence of tobacco, and alcohol consumption data from Global Health Observatory (GHO) data repository from World Health Organization (WHO) [@citeWHO]. We construct a multiple linear regression model in which life expectancy at age 60 for all WHO regions is explained by five variables, sex, region, year, prevalence of tobacco, and alcohol consumption. We find that female has a higher life expectancy at age 60 than male across years, and life expectancy has a large region difference as well as tobacco use and alcohol consumption. We also find as the prevalence of tobacco increases, the expected life expectancy decreases while as the alcohol consumption increases, the expected life expectancy also increases. Further work could introduce more complex models to capture non-linear relationships and give region-specific studies as well as taking into account other socioeconomic and environmental factors.

The remainder of this paper is structured as follows: @sec-data discusses the data with @sec-data_description including information about data collection and data cleaning results, and @sec-data_visualization including graphs and tables representing relationships between variables and some discussions; @sec-model introduces two linear regression models we employ and specifies parameters; @sec-result shows the estimates of fitted models and criteria for comparing two models; @sec-discussion includes discussions about model result and implications with @sec-summary making a brief summary of what the paper has done, @sec-findings showing findings based on model and data visualization, and @sec-weakness talking about the weakness and future improvement.

@@

The datasets used in this paper was obtained from World Health Organization (WHO) Global Health Observatory data repository, and is publicly available from WHO website [@citeWHO]. We utilized 3 datasets: life expectancy and Health life expectancy by WHO region [@citeLifeexp], SDG Target 3.5 Substance abuse by WHO region [@citeAlcohol], and SDG Target 3.a Tobacco control by WHO region [@citeTobacco]. The life expectancy dataset contains life expectancy and healthy life expectancy at birth and at age 60 for different sexes and WHO regions in year 2000, 2010, 2015, and 2019. 
The life expectancy values were estimated based on mortality data from civil registration. The substance abuse dataset contains total alcohol per capita (aged 15+) consumption (total APC) for different sexes and WHO regions in year 2000, 2005, 2010, 2015, and 2019. The tobacco control dataset contains the percentage of the population aged 15+ who currently use tobacco products based on population-based surveys for different sexes and WHO regions in year 2000, 2005, 2010, 2015, 2018, 2019, and 2020.

@@

* region: WHO regions
* year: Year data collected
* sex: Sex at birth
* life_expectancy: The average number of years that a person of age 60 could expect to live
* alcohol_consumption: Total alcohol per capita (aged 15+) consumption in litre
* prevalence_of_tobacco: The percentage of the population aged 15+ who currently use any tobacco products.

@@

According to @tbl-model and @fig-tobacco, when other variables are held constant, the increase of the prevalence of tobacco use will decrease the life expectancy at age 60. This is consistent with the fact that tobacco is a health risk factor for cardiovascular and respiratory diseases and thus people with a habit of use tobacco products not only cigarettes will have a lower life expectancy at age 60[@citeTobaccouse]. Combined with @tbl-year and @tbl-region, we could find that although the prevalence of tobacco in general decreases across years, it has an uneven region distribution and notably for Africa, it has the lowest prevalence of tobacco but relatively low life expectancy.

>>>>

---
title: "Lifestyle Choice Matters: Exploring Population Health Disparities and the Effect of Tobacco and Alcohol on Life Expectancy"
author: "Yufei Liu"
thanks: "Code and data are available at: https://github.com/Florence-Liu/life_expectancy"
date: today
date-format: long
abstract: "Life expectancy is an important measurement of population health and there are many significant influencing factors such as genetic and lifestyle. This paper uses a linear regression model to analyze life expectancy, prevalence of tobacco products, and alcohol consumption data from Global Health Observatory data repository of WHO. We find as prevalence of tobacco increases, the life expectancy at age 60 decreases while as alcohol consumption increases, the life expectancy increases. We also find regional, temporal, and sexual differences in life expectancy at age 60. It implies health disparies in several aspects and encourages the promotion of health lifestyle including moderate drinking and taking less tobacco products that contribute to increased life expectancy. "
format: pdf
number-sections: true
toc: false
bibliography: references.bib
---

@@

In this paper, we investigate how lifestyle factors influence life expectancy at age 60, and focus on two specific aspects, prevalence of tobacco products and alcohol consumption. We use `R` [@citeR] to analyze life expectancy, prevalence of tobacco, and alcohol consumption data from Global Health Observatory (GHO) data repository from World Health Organization (WHO). We construct multiple linear regression models in which life expectancy at age 60 for all WHO regions is explained by five variables, sex, region, year, prevalence of tobacco, and alcohol consumption. We find that female has a higher life expectancy at age 60 than male across years, and life expectancy has a large region difference as well as tobacco use and alcohol consumption. We also find as the prevalence of tobacco increases, the expected life expectancy decreases while as the alcohol consumption increases, the expected life expectancy also increases. Further work could introduce more complex models to capture non-linear relationships and give region-specific studies as well as taking into account other socioeconomic and environmental factors.

The remainder of this paper is structured as follows: @sec-data discusses the data with @sec-data_description including information about data collection and data cleaning results, and @sec-data_visualization including graphs and tables representing relationships between variables and some discussions; @sec-model introduces two linear regression models we employ and specifies parameters; @sec-result shows the estimates of fitted models and criteria for comparing two models; @sec-discussion includes discussions about model results and implications with @sec-summary making a brief summary of what the paper has done, @sec-findings showing findings based on model results and data visualization, and @sec-weakness talking about the weaknesses and future improvement.

@@

The datasets used in this paper was obtained from World Health Organization (WHO) Global Health Observatory data repository, and is publicly available from WHO website. We utilized 3 datasets: life expectancy and Health life expectancy by WHO region [@citeLifeexp], SDG Target 3.5 Substance abuse by WHO region [@citeAlcohol], and SDG Target 3.a Tobacco control by WHO region [@citeTobacco]. The life expectancy dataset contains life expectancy and healthy life expectancy at birth and at age 60 for different sexes and WHO regions in year 2000, 2010, 2015, and 2019. 
The life expectancy values were estimated based on mortality data from civil registration. The substance abuse dataset contains total alcohol per capita (aged 15+) consumption (total APC) for different sexes and WHO regions in year 2000, 2005, 2010, 2015, and 2019. The tobacco control dataset contains the percentage of the population aged 15+ who currently use tobacco products based on population-based surveys for different sexes and WHO regions in year 2000, 2005, 2010, 2015, 2018, 2019, and 2020.

@@

* region: WHO regions
* year: Year data collected
* sex: Sex at birth
* life_expectancy: The average number of years that a person of age 60 could expect to live
* alcohol_consumption: Total alcohol per capita aged 15+ consumption in litre
* prevalence_of_tobacco: The percentage of the population aged 15+ who currently use any tobacco products.

@@

According to @tbl-model and @fig-tobacco, when other variables are held constant, the increase of the prevalence of tobacco use will decrease the life expectancy at age 60. This is consistent with the fact that tobacco is a health risk factor for cardiovascular and respiratory diseases and thus people with a habit of use tobacco products not only cigarettes will have a lower life expectancy at age 60 [@citeTobaccouse]. Combined with @tbl-year and @tbl-region, we could find that although the prevalence of tobacco in general decreases across years, it has an uneven region distribution and notably for Africa, it has the lowest prevalence of tobacco but relatively low life expectancy.
